HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-19 scnn.tex
symmetries of Fourier modes
statistical layer

% MATH SYMBOLS
\newcommand{\1}{\b{1}}              % all-ones vector
\newcommand{\0}{\b{0}}              % all-zero vector
\newcommand{\g}[1]{\b{#1}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\L}{\b{L}}
\newcommand{\tL}{\tilde{\L}}
\newcommand{\W}{\b{W}}
\newcommand{\I}{\b{I}}
\newcommand{\D}{\b{D}}
\newcommand{\U}{\b{U}}
\newcommand{\x}{\b{x}}
\newcommand{\X}{\b{X}}
\newcommand{\y}{\b{y}}
\newcommand{\Y}{\b{Y}}
\newcommand{\bu}{\b{u}}
\newcommand{\f}{\b{f}}
\newcommand{\trans}{^\intercal}
\newcommand{\R}{\mathbb{R}}
\newcommand{\bLambda}{\b{\Lambda}}
\newcommand{\blambda}{\b{\lambda}}
\newcommand{\bO}{\mathcal{O}}
\newcommand{\pkg}[1]{\texttt{#1}}

@@

    \item (TK) make the bibliography display only a few name
    % \item (yes) Michael, do you agree with the title HealpixNet? I (Michaël) extensively modified the abstract to reflect that it is a special case of graph CNNs tailored to Healpix. Hence I now agree with the title. Feel free to edit. Ideally I'd like to have "graph" in the title (graph CNNs are quite hot these days), but I couldn't find how.
    % \item Emphasize the novelty of using it on Healpix in abstract, into and rest of the paper
    \item (TK) add info on comparison with histograms, so that it's just a demonstration
    % \item Finish section on Convolutions on graphs (Michael - Nathanael)
    % \item finish section on Efficient convolutions (Michael)
    % \item (done) Plot of the new filters from Mass maps
    % \item (done!, plot to be added) check the power spectrum after smoothing
    % \item (done!) upgrade the plot of pooling (Nathanael)
    \item (TK) possibly improve plot of pooling
	% \item (done!) new plot for network structure (teaser plot) (Michael) What do you think? Todo: link the points on the sphere to the fully connected \nati{I like the plot, but I am thinking about putting multiple spheres with different maps for the convolutiion instead…}
    \item (TK NP) if we have time make a new candidate for network plot (optional)
    % \item (done!) Figure 1 to be round
    \item (first version done) conclusions
    % \item (done!) new simulations with 3 arcmin smoothing
    % \item (done!) simulation with cross-validation for the histogram
    % \item (done!) Update result section according to the new simulations...
    % \item (done!) simulation PSD
    % \item (done!) Results figure, add sigma8
    % \item (more or less done!) possibly move the eigen decomposition and heat into appendix + add discussion (Nathanael I am not sure about the heat diffusion example...)
    \item Should the heat example go to appendix
    \item TK read the heat section again
    \item TK read the result section
    \item TK check the appendix
    \item TK optional - example of data (CMB (mask) + mass map + radio intensity)
    \item TK read everything
	\item MD add irregular (random) sampling to comparison with spherical harmonics? Could motivate people to think about different samplings.
	\item uniformize text size in figures
\end{itemize}

@@

% cohen2017convolutional (ICML workshop) and cohen2018spherical (ICLR best paper) are redundant
A first attempt to generalize CNNs to spherical data was to apply a standard 2D CNN to a grid discretization of the sphere~\citep{boomsma2017spherical}. While this approach leverages the well-developed 2D convolution and hierarchical pooling, it is restricted to those very specific grid samplings which are highly irregular on the sphere.
A second attempt by~\citet{cohen2018spherical} was to leverage the spherical Fourier transform and to perform the convolution in the spectral domain, thanks to the convolution theorem. While there exists fast spherical Fourier transforms for some samplings, this approach is still computationally expensive (see \figref{filtering_speed}).
Finally, the use of a graph to model the discretized sphere was considered by~\citet{khasanova2017graph} as well, towards the goal of building invariances in the network. While their architecture is fully invariant to translation and rotation on the sphere, they did not take advantage of a hierarchical pixelization of the sphere, and the statistical layer used to be invariant needs to be hand-crafted to capture statistics of interest.

@@

The use of a graph to model the discretized sphere was considered by~\citet{khasanova2017graph} as well. The authors map omnidirectional images to the sphere, with the goal of correcting for distortions induced by the lens used for the acquisition.
Besides the different application covered by the paper, there are two important
differences. First the sampling is different as~\cite{khasanova2017graph} uses a
grid. Second, it uses a different graph CNN based on a different parameterization
of the convolution kernel.
As~\citep{boomsma2017spherical}, they use an equirectangular grid sampling. Moreover, they did not take advantage of a hierarchical pixelization of the sphere and resorted to dynamic pooling~\citep{kalchbrenner2014dcnn}.
Finally, while we do want our network to be rotation invariant, their statistical layer --- an operation which computes a set of spatially invariant statistics of the last feature map --- destroys a lot of information \todo{and hinders performance (should be empirically verified)}. Moreover, the chosen statistics have to be hand-crafted to capture statistics of interest. While developing completely invariant transforms is a  goal worth investigating, CNNs for image recognition (arguably the most researched and developed architectures), are not built to be fully invariant to translation. As with the proposed spherical CNN, the fully connected layers at the end of the network depend on the spatial localization of features. While that may not be ideal, it was shown to work well in practice.
\todo{The statistical layer, an idea that we reused in this work. See xxx.}

@@

% The advantage of schemes based on a base polyhedron is that cells are equal-area. While that is not required for the convolution operation when modeling with graphs (it can be corrected by setting edge weights appropriately, see \citet{khasanova2017graph})
% nope: can also do a regular pooling on the equirectangular. It preserves the sampling, and the convolution is adapted to the sampling thanks to the edge weights.

@@

@@

A typical neural network is constructed as a stack of layers, which sequentially transform the input data. Our architecture, pictured in \figref{architecture}, is composed of the following types of layers: graph convolutional (GC) layers, non-linearity, pooling (P), batch normalization (BN), and fully connected (FC) layers. All GC and FC layers are always followed by a non-linearity. As such, we will not mention the non-linearity in the description of our architectures.

@@

@@

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{subspace_harmonics_eigenvectors}
	\caption{Correspondence between the subspaces spanned by the graph Fourier modes and the spherical harmonics. Each pixel on the plot is the dot product between one spherical harmonic and one graph Fourier mode $\b u_i$. The plot clearly shows how the subspaces are aligned: modes and harmonics which correspond to different degrees $\ell$ are mostly orthogonal to each other, while modes and harmonics which correspond to the same degree $\ell$ are linearly dependent. There is $2 \ell + 1$ orders per degree $\ell$.}
	\label{fig:subspace_harmonics_eigenvectors}
\end{figure}

>>>>

% MATH SYMBOLS
\newcommand{\1}{\b{1}}              % all-ones vector
\newcommand{\0}{\b{0}}              % all-zero vector
\newcommand{\g}[1]{\b{#1}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\E}{\mathcal{E}}
\renewcommand{\L}{\b{L}}
\newcommand{\tL}{\tilde{\L}}
\newcommand{\W}{\b{W}}
\newcommand{\I}{\b{I}}
\newcommand{\D}{\b{D}}
\newcommand{\U}{\b{U}}
\newcommand{\x}{\b{x}}
\newcommand{\X}{\b{X}}
\newcommand{\y}{\b{y}}
\newcommand{\Y}{\b{Y}}
\newcommand{\bu}{\b{u}}
\newcommand{\f}{\b{f}}
\newcommand{\trans}{^\intercal}
\newcommand{\R}{\mathbb{R}}
\newcommand{\bLambda}{\b{\Lambda}}
\newcommand{\blambda}{\b{\lambda}}
\newcommand{\bO}{\mathcal{O}}
\DeclareMathOperator*{\esp}{E}
\DeclareMathOperator*{\var}{Var}
\newcommand{\pkg}[1]{\texttt{#1}}

@@

    \item (TK) make the bibliography display only a few name
	\item check if papers which appear as arxiv were not published in a conference or journal
    % \item (yes) Michael, do you agree with the title HealpixNet? I (Michaël) extensively modified the abstract to reflect that it is a special case of graph CNNs tailored to Healpix. Hence I now agree with the title. Feel free to edit. Ideally I'd like to have "graph" in the title (graph CNNs are quite hot these days), but I couldn't find how.
    % \item Emphasize the novelty of using it on Healpix in abstract, into and rest of the paper
    \item (TK) add info on comparison with histograms, so that it's just a demonstration
    % \item Finish section on Convolutions on graphs (Michael - Nathanael)
    % \item finish section on Efficient convolutions (Michael)
    % \item (done) Plot of the new filters from Mass maps
    % \item (done!, plot to be added) check the power spectrum after smoothing
    % \item (done!) upgrade the plot of pooling (Nathanael)
    \item (TK) possibly improve plot of pooling
	% \item (done!) new plot for network structure (teaser plot) (Michael) What do you think? Todo: link the points on the sphere to the fully connected \nati{I like the plot, but I am thinking about putting multiple spheres with different maps for the convolutiion instead…}
    \item (TK NP) if we have time make a new candidate for network plot (optional)
    % \item (done!) Figure 1 to be round
    \item (first version done) conclusions
    % \item (done!) new simulations with 3 arcmin smoothing
    % \item (done!) simulation with cross-validation for the histogram
    % \item (done!) Update result section according to the new simulations...
    % \item (done!) simulation PSD
    % \item (done!) Results figure, add sigma8
    % \item (more or less done!) possibly move the eigen decomposition and heat into appendix + add discussion (Nathanael I am not sure about the heat diffusion example...)
    \item Should the heat example go to appendix
    \item TK read the heat section again
    \item TK read the result section
    \item TK check the appendix
    \item TK optional - example of data (CMB (mask) + mass map + radio intensity)
    \item TK read everything
	\item MD add irregular (random) sampling to comparison with spherical harmonics? Could motivate people to think about different samplings.
	\item uniformize text size in figures
\end{itemize}

@@

% cohen2017convolutional (ICML workshop) and cohen2018spherical (ICLR best paper) are redundant
A first attempt to generalize CNNs to spherical data was to apply a standard 2D CNN to a grid discretization of the sphere~\citep{boomsma2017spherical}. While this approach leverages the well-developed 2D convolution and hierarchical pooling, it is restricted to those very specific grid samplings which are highly irregular on the sphere.
A second attempt by~\citet{cohen2018spherical} was to leverage the spherical Fourier transform and to perform the convolution in the spectral domain, thanks to the convolution theorem. While there exists fast spherical Fourier transforms for some samplings, this approach is still computationally expensive (see \figref{filtering_speed}).
Finally, the use of a graph to model the discretized sphere was considered by~\citet{khasanova2017graphomni} as well, towards the goal of building invariances in the network. While their architecture is fully invariant to translation and rotation on the sphere, they did not take advantage of a hierarchical pixelization of the sphere, and the statistical layer used to be invariant needs to be hand-crafted to capture statistics of interest.

@@

The use of a graph to model the discretized sphere was considered by~\citet{khasanova2017graphomni} as well. The authors map omnidirectional images to the sphere, with the goal of correcting for distortions induced by the lens used for the acquisition.
Besides the different application covered by the paper, there are two important
differences. First the sampling is different as~\cite{khasanova2017graphomni} uses a
grid. Second, it uses a different graph CNN based on a different parameterization
of the convolution kernel.
As~\citep{boomsma2017spherical}, they use an equirectangular grid sampling. Moreover, they did not take advantage of a hierarchical pixelization of the sphere and resorted to dynamic pooling~\citep{kalchbrenner2014dcnn}.
Finally, while we do want our network to be rotation invariant, their statistical layer --- an operation which computes a set of spatially invariant statistics of the last feature map --- destroys a lot of information \todo{and hinders performance (should be empirically verified)}. Moreover, the chosen statistics have to be hand-crafted to capture statistics of interest. While developing completely invariant transforms is a  goal worth investigating, CNNs for image recognition (arguably the most researched and developed architectures), are not built to be fully invariant to translation. As with the proposed spherical CNN, the fully connected layers at the end of the network depend on the spatial localization of features. While that may not be ideal, it was shown to work well in practice.
\todo{In this work we reused the idea of a statistical layer, described in \secref{stat_layer}, to provide invariance to isometric transformations such as translation and rotation on the sphere. While \citet{khasanova2017graphomni} uses the mean and variance of diffused versions of the last feature maps, we experimentally tested multiple variants and got good results by keeping all the statistical moments via a learned histogram. That is similar to the learnable histogram introduced by \citet{wang2016learnhist}, although for the orthogonal purpose of integrating global context in object detection and segmentation tasks.}

@@

% The advantage of schemes based on a base polyhedron is that cells are equal-area. While that is not required for the convolution operation when modeling with graphs (it can be corrected by setting edge weights appropriately, see \citet{khasanova2017graphomni})
% nope: can also do a regular pooling on the equirectangular. It preserves the sampling, and the convolution is adapted to the sampling thanks to the edge weights.

@@

\subsection{Statistical layer}
\label{sec:stat_layer}
% Or Global pooling?

Following \citet{khasanova2017tigranet}, we use a statistical layer between the last graph convolutional layer and the first fully connected layer for the neural network to be invariant to isometric transformations such as rotation and translation of maps on the sphere. The role of the layer is to compute global and location independent statistics from feature maps. Besides providing invariance, the layer drastically compresses the representation of the data as the number of computed statistics $N_{stat}$ is usually much smaller than the size of the map, $N_{pix}$. That compression results in much smaller fully connected layers, the most parameter-hungry layers of a neural network. The invariance and reduced representation should help the network to learn from less samples and to generalize better to unseen data. Yet another advantage is that a network built with such a statistical layer is agnostic to the size of the input data, i.e., it can classify maps which do not have the same number of pixels. Indeed, such global summarization is invariant to node permutation and is commonly used along graph convolutions to classify graphs of varying sizes \citep{duvenaud2015gcn, li2015gatedgnn}.

Given a feature map $\x \in \R^{N_{pix}}$, the statistical layer is defined as
\begin{equation} \label{eqn:stat_layer}
	\y = f(\x) \in \R^{N_{stat}},
\end{equation}
where $f$ computes $N_{stat}$ statistics which are independent from the size of the map, $N_{pix}$. An example is the global average pooling $f(\x) = \esp(\x)$, introduced by \citet{lin2013globalavgpooling}, where $\esp(\x) = \frac{1}{N_{pix}} \sum_{i=1}^{N_{pix}} x_i$ is the empirical mean. Another example is $f(\x) = [\esp(\x), \var(\x)]\trans$, where $\var(\x) = \frac{1}{N_{pix}} \sum_{i=1}^{N_{pix}} (x_i - \esp(\x))^2$ is the empirical variance.

\todo{learned histogram \citep{wang2016learnhist} if it shows good results}

\todo{stats of diffused maps \citep{khasanova2017tigranet}. Provides stats at multiple scales?}

\todo{perf w.r.t. different functions $f$. Here or in the experiment section?}

@@

Neural networks are constructed as stacks of layers which sequentially transform the input data. Our architecture, pictured in \figref{architecture}, is composed of the following types of layers: graph convolutional (GC) layers, non-linearity, pooling (P), batch normalization (BN), statistical layer (ST), and fully connected (FC) layers. All GC and FC layers are always followed by a non-linearity. As such, we will not mention the non-linearity in the description of our architectures.

@@

The ST layer, described by \eqnref{stat_layer}, independently transforms each of the $F$ feature maps of the input $\X \in \R^{N_{pix} \times F}$ into $\Y \in \R^{N_{stat} \times F}$, where $N_{stat}$ is the number of statistics. One may even omit the FC layers and directly fed the output of the ST layer into the softmax layer, as proposed by \citet{lin2013globalavgpooling}. In this regime, $N_{stat} = 1$ and $F$ should be equal to the number of classes to discriminate, i.e., the last GC layer outputs as many feature maps as there is classes.

@@

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{subspace_harmonics_eigenvectors}
	\caption{Correspondence between the subspaces spanned by the graph Fourier modes and the spherical harmonics. Each pixel on the plot is the dot product between one spherical harmonic and one graph Fourier mode $\b u_i$. The plot clearly shows how the subspaces are aligned: modes and harmonics which correspond to different degrees $\ell$ are mostly orthogonal to each other, while modes and harmonics which correspond to the same degree $\ell$ are linearly dependent. There is $2 \ell + 1$ orders per degree $\ell$. Moreover, we see that the Fourier modes have the correct symmetries, indicated by the perfect orthogonality between even and odd functions.}
	\label{fig:subspace_harmonics_eigenvectors}
\end{figure}
